Efficient Data Provenance Analysis for Concurrent Programs

### 2013-11-13 paper.tex
committing more changes.  Up through page 6 or so.

\section{Introduction}
Multi-threaded programming is challenging.  In contrast to 
sequential reasoning, multi-threaded programs require complex reasoning about
many threads of control and their interactions.  In shared memory programs,
threads interact by reading and writing memory locations.  Threads' reads and
writes interleave arbitrarily and nondeterministically and each interleaving
can yield different behavior.  Reasoning about thread interactions is a central
challenge to building multi-threaded software because some thread interactions
have undesirable consequences, like causing a crash or degrading performance.

@@

These efforts have shown some of the promise of provenance, but are not a
debugging panacea and are insufficient for dealing with concurrent program
bugs. First, prior work~\cite{badapples} has focused on providing provenance
information for only some unusable {\em values}, like {\tt NULL}, which limits its applicability. 
% Limiting the applicability of their technique to only some failures, like {\tt
% Null Pointer Exceptions}.
Second, while prior work has shown that limited provenance tracking for Java
programs is possible with overheads low enough for production~\cite{badapples},
a more general form of provenance tracking for C and C++ programs has been
elusive. Prior approaches that build on slicing~\cite{thinslicing,
tipslicingsurvey, whylineicse} or heavy-weight memory checkers~\cite{badapples}
impose overheads too high for production for C and C++ programs. Third, no
prior technique that we know of has collected provenance information that is
tailored to debugging concurrent programs with overheads that are low enough
for use in production systems. Addressing concurrency is important because a
value could have originated in any of a large number of threads, making
reasoning about why a variable contains that value complex and difficult. In
summary, the challenges in using provenance to debug concurrent programs are:
(1) collecting the right information to characterize the concurrent execution
behavior; and (2) making it efficient enough for general programs running in
production settings.

In this work we address the challenges above and develop a novel form of
in-production provenance tracking for concurrent programs. We have three
goals. (1) To collect a general form of provenance information, not limited to
only certain values. (2) To provide provenance information that helps
understand how threads interact, which is the key to understanding concurrency
bugs. (3) To work with overheads low enough for use in production (ideally,
less than 10\%).

We develop a novel provenance tracking mechanism called {\em last writer
slicing} that satisfies these goals.   As the program executes in production,
last writer slicing tracks the thread and program point that last wrote any
value to each potentially shared memory location --- its provenance.  Last
writer slices are saved with a core dump when the program fails and are
available in the debugger ({\em e.g.}, for breakpoint debugging).

Last writer slices are primarily useful for concurrency debugging, but we also
observe that provenance information in the form of last writer slices is a
useful foundation for implementing other concurrent program analyses.  Several
tools from prior work has used {\em ad hoc} last-writer information as part of
more heavy-weight analyses in domains like bug detection ({\em e.g.},
~\cite{dmtracker,avio}), automatic debugging ({\em e.g.},
~\cite{recon,conseq,cci}), profiling ({\em
e.g.},~\cite{threadclustering, schedpredictionmodel}), and program
understanding~\cite{oshatr}.

@@

\item{We describe last writer slicing, an novel data provenance tracking
mechanism, that produces information that is useful for debugging concurrent
programs.  We design and implement system support for collecting last writer
slices that is efficient enough for use in production.  }

@@

Data provenance information is directly useful during program debugging.  prior
work~\cite{whylinechi,whylineicse} showed that when programmers debug
programs, they benefit from the ability to directly ask such provenance
(``why'') questions of their debugger.  Origin tracking~\cite{badapples} and
program slicing~\cite{tipslicingsurvey} also make debugging easier by reporting
the provenance of values.  Like these prior efforts, in this work we focus on
using data provenance information for {\em debugging} -- understanding the root
cause of a failure -- as opposed to {\em bug detection} -- finding new, unknown
bugs.

@@

These analyses vary in their purpose and environments, but a cross-cutting
theme is the advantage reaped from their use in a deployment environment.
Debugging tools, profilers, and anomaly detectors benefit from seeing diverse,
real-world behavior.  Dynamic failure avoidance techniques must work in
deployed systems to be effective.  Deployment environments demand low overheads
and all of these analyses benefit from a general, high-performance
communication tracking framework.  Unfortunately, many of these techniques are implemented using heavy-weight infrastructure, like binary instrumentation~\cite{pin}, making them inappropriate for use in production.

@@

Figure~\ref{fig:basicLWT} shows the basic operation of the \lwt.  The right
side of the figure shows a program execution with three threads.  On the left
is the \lwt at the point of each memory access.  There are two key things to
notice.  First, write operations update the \lwt entry of the memory location
they access.  Second, there is no action necessary for read operations.
Making read operations cheap is key to making the overhead of last writer
slicing low enough for deployment use.

@@

At each point in an execution, a memory location's \lwt entry represents the
provenance of the data stored in that memory location.   Last writer slicing
performs no action on read operations and to be useful, it does not need to.
Instead, a programmer can examine an \lwt entry at any arbitrary point in an
execution ({\em e.g.}, at a breakpoint), including after a failure, and see a
value's provenance.  Including the \lwt with core dumps means provenance is
available when post-mortem debugging production failures.

@@

The \lwt implemented in the runtime library is a {\em shadow memory} that
holds one entry for each potentially shared memory address accessed by the
program.  Each \lwt entry records the thread and program point that last wrote
to the entry's corresponding memory location.  When a thread performs a write
to a memory location, it updates the \lwt by recording its thread identifier
and the program point of its write operation to the memory location's \lwt
entry.  A program point may be an instruction address, or a call stack.  We use
instructions by default, but we experimented with 2-address bounded call-stacks
as well.

Collecting last writer slices requires instrumentation on each write operation.
We built an instrumenting compiler pass that inserts a call to an \lwt update
function in our runtime library just before each write that may access data
last written by another thread (escaping memory accesses).  Write
instrumentation calls are passed three pieces of information: the address of
the memory word involved in the access, the thread identifier of the accessing
thread, and the accessing program point.   We use a {\tt GCC}'s local function
escape analysis to identify accesses to memory locations that are provably
inaccessible to code running in other threads.  Such operations access unshared
data and they are not instrumented by our compiler.

@@

For our design to be correct, $LWT[m]$, the entry for a memory location $m$ in
our \lwt implementation, must be the same as $LWS[m]$ at all points in an
execution.  Correctness in the \lwt requires {\em soundness} and {\em
completeness}.  A {\em sound} design only updates $LWT[m]$ when $m$ is written.
A {\em complete} design always updates $LWT[m]$ when $m$ is written.  Our
compiler inserts \lwt updates before every write operation (for completeness)
and only before write operations (for soundness).  Formally, given a program,
$P$, our compiler pass transforms it to a new program, $P'$.  Each thread in
$P'$ executes a modified sequence of instructions: $P'_{t} \in P' = \{
I(p^{t}_{1}), p^{t}_{1}, \ldots, I(p^{t}_{k_{t}}), p^{t}_{k_{t}} \}$.
$I(p^{t}_{j})$ sets $LWT[m] = (t,p^{t}_{j})$ if and only if $p^{t}_{j} =
Write(m)$ and is a no-op otherwise.  This reasoning shows our \lwt is both
sound, complete, and thus correct.

@@

\paragraph{Runtime Support}
The \ctraps runtime implements the trap hook function,  which has two purposes.
The trap hook function detects communication and delivers traps to \ctraps
applications when communication occurs.  The trap hook function detects
communication by looking up the \lwt entry for the memory location being
accessed.  If the thread identifier in the \lwt entry is different from the
accessing thread, then the trap hook function delivers a trap to the accessing
thread, just before the memory access is allowed to proceed.  To deliver a
trap, the trap hook function calls {\em trap handler routines} that are defined
in \ctraps applications.  Trap handler routines are like signal handlers.  The
\ctraps runtime maintains a list of trap handlers that are registered when the
program starts via a configuration stored in the environment.  Trap handlers
may contain arbitrary application-specific code.  \ctraps defines a trap
handling API.  Trap handlers are passed the four pieces of information passed
to the trap hook, as well as the \lwt's record of the program point and thread
identifier of the last write to the involved memory location.

@@

We implemented last writer slicing and \ctraps.  We built our compiler support
as a plugin for GCC ({\tt gcc-4.7.0 20111112}).  We used GCC's points-to and
escape analysis support to prune instrumentation to non-escaping memory
locations.  Our compiler instruments calls to {\tt free} and {\tt delete} as
writes to the pointer being deallocated.  Our analysis handles all code
compilable using this version of GCC except for a few cases: we do not handle
accesses that compile to GCC {\tt BITFIELD\_REFERENCE} IR types because these
are undocumented, we do not handle inlined assembly instructions, and we do not
handle C++ exceptions.  Note that these limitations of our research prototype
are not fundamental and the limited cases are uncommon.

@@

\section{Evaluation}
\label{sec:eval}
We evaluated last writer slices and \ctraps along several axes.  We show that
last writer slices provide information that is useful during debugging in
Section~\ref{sec:eval:debugging} looking at a set of buggy programs and
comparing to prior work.  We evaluate our performance overheads and
show that last writer slices and \ctraps are feasible for use in deployed
systems for a majority of workloads we studied.  We then characterize our
instrumentation and discuss sources run time overhead.   We
show that \ctraps is useful by implementing two existing applications and
analyzing their performance.

@@

In Table~\ref{tab:bugs}, we show the bugs we studied describing the program,
bug type, and bug report identifier where applicable.  For each bug, we
followed the procedure for installation and bug triggering described by the
author of~\cite{concurrencybugs}.  We study a mixture of concurrency bug types
in this work, including both single- and multi-variable bugs and both atomicity
and ordering violations.

The table also summarizes our evaluation of last writer slicing's debugging
benefits.  We mirror the evaluation strategy of bad value origin
tracking~\cite{badapples}, as it is the only other low-overhead provenance
tracking mechanism we know of.  We evaluate our system using their three
evaluation criteria - (1) {\bf effectiveness}, (2) {\bf triviality}, and (3)
{\bf usefulness}.  Our technique is {\bf effective} if the last writer slice
reports a code point that wrote a value that subsequently caused a failure.  A
failure is {\bf trivial} to debug if its cause is made obvious by examining the
core dump alone, {\em i.e.}, a last writer slice is unnecessary.  A last writer
slice is {\bf useful} if it is effective and the thread and code point in the
slice help understand why the failure occurred or how to fix it.   We directly
compare to bad value origin tracking.    The {\bf O.T.} column in
Table~\ref{tab:bugs} shows  whether their analysis could help debug each bug,
based on the description of their technique.

>>>>

\section{Introduction}
Multi-threaded programming is challenging.  In contrast to 
sequential reasoning, multi-threaded programs require complex reasoning about
many threads of control and their interactions.  In shared memory programs,
threads interact by reading and writing memory locations.  Threads' reads and
writes interleave arbitrarily and nondeterministically and each interleaving
can yield different behavior.  Reasoning about thread interactions is a central
challenge to building multi-threaded software because some thread interactions
have undesirable consequences, like crashing or degrading performance.

@@

These efforts have shown some of the promise of provenance, but are not a
debugging panacea and are insufficient for dealing with concurrent program
bugs. First, prior work~\cite{badapples} has focused on providing provenance
information for only some unusable {\em values}, like {\tt NULL}, which limits its applicability. 
% Limiting the applicability of their technique to only some failures, like {\tt
% Null Pointer Exceptions}.
Second, while prior work has shown that limited provenance tracking for Java
programs is possible with overheads low enough for production~\cite{badapples},
a more general form of provenance tracking for C and C++ programs has been
elusive. Prior approaches that build on slicing~\cite{thinslicing,
tipslicingsurvey, whylineicse} or heavy-weight memory checkers~\cite{badapples}
impose overheads too high for production for C and C++ programs. Third, no
prior technique that we know of has collected provenance information that is
tailored to debugging concurrent programs with overheads that are low enough
for use in production systems. Addressing concurrency is important because a
value could have originated in any of a large number of threads, making
reasoning about why a variable contains that value complex and difficult.

%In
%summary, the challenges in using provenance to debug concurrent programs are:
%(1) collecting the right information to characterize the concurrent execution
%behavior; and (2) making it efficient enough for general programs running in
%production settings.

In this work we address the challenges above and develop a novel form of
in-production provenance tracking for concurrent programs. We have three goals.
(1) To collect a general form of provenance information, not limited to only
certain values. (2) To provide provenance information that helps understand how
threads interact, which is the key to understanding concurrency bugs. (3) To
work with overheads low enough for use in production (ideally, less than 10\%),
especially for long-running server programs that experience a diversity of production behavior.

We develop a new provenance tracking mechanism called {\em last writer
slicing} that satisfies these goals.   As the program executes in production,
last writer slicing tracks the thread and program point that last wrote any
value to each potentially shared memory location --- its provenance.  Last
writer slices are saved with a core dump when the program fails and are
available in the debugger ({\em e.g.}, for breakpoint debugging).

Last writer slices are primarily useful for concurrency debugging, but we 
observe that provenance information is a
useful foundation for implementing other concurrent program analyses.  Several
tools from prior work have used {\em ad hoc} last-writer information as part of
more heavy-weight analyses in domains like bug detection ({\em e.g.},
~\cite{dmtracker,avio}), automatic debugging ({\em e.g.},
~\cite{recon,conseq,cci}), profiling ({\em
e.g.},~\cite{threadclustering, schedpredictionmodel}), and program
understanding~\cite{oshatr}.

@@

\item{We describe last writer slicing, a novel data provenance tracking
mechanism, that produces information that is useful for debugging concurrent
programs.  We design and implement system support for collecting last writer
slices that is efficient enough for use in production.  }

@@

Data provenance information is directly useful during program debugging.  Prior
work~\cite{whylinechi,whylineicse} showed that when programmers debug
programs, they benefit from the ability to directly ask such provenance
(``why'') questions of their debugger.  Origin tracking~\cite{badapples} and
program slicing~\cite{tipslicingsurvey} also make debugging easier by reporting
the provenance of values.  Like these prior efforts, in this work we focus on
using data provenance information for {\em debugging} -- understanding the root
cause of a failure -- as opposed to {\em bug detection} -- finding new, unknown
bugs.

@@

These analyses vary in their purpose and environments, but a cross-cutting
theme is the advantage reaped from their use in a deployment environment.
Debugging tools, profilers, and anomaly detectors benefit from seeing diverse,
real-world behavior.  Dynamic failure avoidance techniques must work in
deployed systems to be effective.  Deployment environments demand low overheads
and all of these analyses benefit from a general, high-performance
communication tracking framework.  Unfortunately, many of these techniques are built using heavy-weight infrastructure, like binary instrumentation~\cite{pin}, with overheads too high for production.

@@

Figure~\ref{fig:basicLWT} shows the basic operation of the \lwt.  The right
side of the figure shows a program execution with three threads.  On the left
is the \lwt at the point of each memory access.  There are three key things to
notice.  First, write operations update the \lwt entry of the memory location
they access.  Second, there is no action necessary for read operations.  Making
read operations cheap is key to making the overhead of last writer slicing low
enough for production use.  Third, if the program stops at a point like the
{\tt Rd Y} illustrated in Thread 1, the \lwt holds provenance information for
the involved values ({\em i.e.}, Thread 3 last wrote {\tt Y} at the code point
shown). That information helps understand {\em why} {\tt Y} holds its value.

@@

At each point in an execution, a memory location's \lwt entry represents the
provenance of the data stored in that memory location.   Last writer slicing
performs no action on read operations and to be useful, it does not need to.
Instead, a programmer can examine an \lwt entry at any arbitrary point in an
execution and see a value's
provenance ({\em e.g.}, at a breakpoint or after a failure).  Including the \lwt with core dumps means provenance is available
when post-mortem debugging production failures.

@@

The \lwt implemented in the runtime library is a {\em shadow memory} that holds
one entry for each potentially shared memory address accessed by the program.
Each \lwt entry records the thread and program point that last wrote to the
entry's corresponding memory location.  When a thread performs a write to a
memory location, our runtime library's \lwt update function updates the \lwt by
recording the thread's identifier and the program point of the write operation
in the memory location's \lwt entry.  A program point may be an instruction
address or a call stack.  We use instructions by default, but we also experimented
with 2-address call stacks.

Collecting last writer slices requires instrumentation on each write operation.
We built an instrumenting compiler pass that inserts a call to the \lwt update
function in our runtime library just before each write that may access data
last written by another thread (escaping memory accesses).  \lwt update 
calls are passed three pieces of information: the address of
the memory word involved in the access, the thread identifier of the accessing
thread, and the accessing program point.   We use a {\tt GCC}'s local function
escape analysis to identify accesses to memory locations that are provably
inaccessible to code running in other threads.  Such operations access unshared
data and we do not instrument them.

@@

For our design to be correct, $LWT[m]$, the entry for a memory location $m$ in
our \lwt implementation, must be the same as $LWS[m]$ at all points in an
execution.  Correctness in the \lwt requires {\em soundness} and {\em
completeness}.  A {\em sound} design only updates $LWT[m]$ when $m$ is written.
A {\em complete} design always updates $LWT[m]$ when $m$ is written.  Our
compiler inserts \lwt updates before every write operation (for completeness)
and only before write operations (for soundness).  Formally, given a program,
$P$, our compiler pass transforms it to a new program, $P'$.  Each thread in
$P'$ executes a modified sequence of instructions: $P'_{t} \in P' = ( 
I(p^{t}_{1}), p^{t}_{1}, \ldots, I(p^{t}_{k_{t}}), p^{t}_{k_{t}} )$.
$I(p^{t}_{j})$ sets $LWT[m] = (t,p^{t}_{j})$ if and only if $p^{t}_{j} =
Write(m)$ and is a no-op otherwise.  This reasoning shows our \lwt is both
sound, complete, and thus correct.

@@

\paragraph{Runtime Support}
The \ctraps runtime implements the trap hook function,  which has two purposes.
The trap hook function detects communication and delivers traps to \ctraps
applications when communication occurs.  The trap hook function detects
communication by looking up the \lwt entry for the memory location being
accessed.  If the thread identifier in the \lwt entry is different from the
accessing thread, then the trap hook function delivers a trap to the accessing
thread, just before the memory access is allowed to proceed.  To deliver a
trap, the trap hook function calls {\em trap handler routines} that are defined
in \ctraps applications.  Trap handler routines are like signal handlers.  The
\ctraps runtime maintains a list of trap handlers that are registered when the
program starts via a configuration stored in the environment.  Trap handlers
may contain arbitrary application-specific code.  \ctraps defines a trap
handling API.  Trap handlers are passed the four pieces of information passed
to the trap hook, as well as the \lwt's record of the program point and thread
identifier of the last write to the involved location.

@@

We implemented last writer slicing and \ctraps.  We built our compiler support
as a plugin for GCC ({\tt gcc-4.7.0 20111112}).  We used GCC's points-to and
escape analysis support to prune instrumentation on accesses to non-escaping memory
locations.  Our compiler instruments calls to {\tt free} and {\tt delete} as
writes to the pointer being deallocated.  Our analysis handles all code
compilable using this version of GCC except for a few cases: we do not handle
accesses that compile to GCC {\tt BITFIELD\_REFERENCE} IR types because these
are undocumented, we do not handle inlined assembly instructions, and we do not
handle C++ exceptions.  Note that these limitations of our research prototype
are not fundamental and the limited cases are uncommon.

@@

\section{Evaluation}
\label{sec:eval}
We evaluated last writer slices and \ctraps along several axes.  We show that
last writer slices provide information that is useful during debugging in
Section~\ref{sec:eval:debugging}, looking at a set of buggy programs and
comparing to prior work.  We evaluate our performance overheads and
show that last writer slices and \ctraps are feasible for use in deployed
systems for a majority of workloads we studied.  We then characterize our
instrumentation and discuss sources of run time overhead.   We
show that \ctraps is useful by implementing two existing applications and
analyzing their performance.

@@

In Table~\ref{tab:bugs}, we show the bugs we studied, describing the program,
bug type, and bug report identifier where applicable.  For each bug, we
followed the procedure for installation and bug triggering described by the
author of~\cite{concurrencybugs}.  We study a mixture of concurrency bug types
in this work, including both single- and multi-variable bugs and both atomicity
and ordering violations.

The table also summarizes our evaluation of last writer slicing's debugging
benefits.  We mirror the evaluation strategy of bad value origin
tracking~\cite{badapples}, as it is the only other low-overhead provenance
tracking mechanism we know of.  We evaluate our system using their three
evaluation criteria - (1) {\bf effectiveness}, (2) {\bf triviality}, and (3)
{\bf usefulness}.  Our technique is {\bf effective} if it tracks a correct last
writer slice for the value or values involved in causing the failure.  A failure is {\bf
trivial} to debug if its cause is made obvious by examining the core dump
alone, {\em i.e.}, a last writer slice is unnecessary.  A last writer slice is
{\bf useful} if it is effective and the thread and code point in the slice help
understand why the failure occurred or how to fix it.   We directly compare to
bad value origin tracking.    The {\bf O.T.} column in Table~\ref{tab:bugs}
shows  whether their analysis could help debug each bug, based on the
description of their technique.
